---
title: "Assignment B-1: Making a function"
output: github_document
---

## Exercise 1 & 2: Make a Function (25 & 20 points)

### `check_outliers()`
I'm going to create a function that checks a numeric vector for any outlier values. An outlier is typically considered to be more than 3 standard deviations from the mean, or 1.5 times greater than the interquartile range. Instead of having to plot a box plot or checking manually, this function will automatically return all the outlier values in a vector. There are options to choose between methods, and set custom thresholds or quantile calculation algorithm if needed.

```{r}

# Making the function -----------------------------------------------------------------------------------------------------------------

check_outliers <- function(x, method, zscore_threshold = 3, iqr_threshold = 1.5, quantile_type = 7) {
  
  # Remove all NA values. We are only calculating and returning outliers, and including NAs would disrupt our mean and SD calculations
  x <- na.omit(x)
  
  # Check if the input x is a numeric vector greater than length 0
  if (!is.numeric(x) || length(x) == 0) {
    stop('This function only works for a numeric vector greater than length 0, please provide the correct class and/or length')
    }


  # Z-score method: calculates z-score using the standard score formula (x-mean/SD) and indexes the vector for outliers outside the threshold
  if (tolower(method) == 'zscore') {
    
    # Validate the z-score threshold
    if (!is.numeric(zscore_threshold) || zscore_threshold < 1) {
      stop("The z-score threshold must be a numeric value at least 1 SD or greater.")
    }
    
    # Calculate mean and SD
    mean_x <- mean(x)
    sd_x <- sd(x)
    
    # Before continuing, check for SD = 0 or NA because then we cannot do the calculations below
    if (sd_x == 0 || is.na(sd_x)) {
      message("Standard deviation is zero or NA; no outliers can be determined using this method.")
      return(numeric(0))  # Return an empty vector to keep output consistent
    }
    
    zscores <- (x - mean_x)/sd_x # Calculate z-scores
    outliers <- x[abs(zscores) > zscore_threshold] # only include values above the threshold
  }
  
  # IQR method: calculates IQR using the inputting quantile algorithm and indexes the vector for outliers outside the threshold
  else if (tolower(method) == 'iqr') {
    
    # Validate the IQR threshold
    if (!is.numeric(iqr_threshold) || iqr_threshold <= 0) {
      stop("The IQR threshold must be a numeric value greater than or equal to 0.")
    }
    
    # Validate the quantile algorithm type
    if (!quantile_type %in% 1:9) {
    stop("Invalid quantile algorithm type; must be an integer between 1 and 9.")
}
    
    # Calculate first and third quartiles, and IQR, using inputted algorithm type
    q1 <- quantile(x, prob = 0.25, type = quantile_type)
    q3 <- quantile(x, prob = 0.75, type = quantile_type)
    iqr_x <- q3 - q1
    
    # Calculate upper and lower bounds
    upper_bound <- q3 + (iqr_threshold * iqr_x)
    lower_bound <- q1 - (iqr_threshold * iqr_x)
    outliers <- x[x > upper_bound | x < lower_bound] # only include values outside these bounds
  }
  
  # If incorrect method input
  else{
    stop('The only supported methods are "zscore" or "iqr" [case insensitive]. Please check for typos and try again')
  }
  
  # Check if any outliers were found; print a message if not
   if (length(outliers) == 0) {
    message("No outliers found.")
     return(numeric(0))
  }
  
  # Return the outliers in a numeric vector
  return(outliers)
}

# Documenting the function ------------------------------------------------------------------------------------------------------------

#' Outlier Detection in Numeric Vectors
#'
#' This function checks for outlier values in a numeric vector based on either the Z-score method or the Interquartile Range (IQR) method. It returns a numeric vector containing the detected outliers, or an empty numeric vector with a message if no outliers were found or an error occurred. It provides flexibility in defining thresholds for outliers and the type of algorithm for calculating quartiles.
#'
#' @param x A numeric vector. This is the primary input for which outliers will be calculated. NA values are automatically omitted within the function to ensure accurate calculations. Named "x" because it is the standard name for a variable in statistics.
#' @param method A string specifying the method for detecting outliers. Options are "zscore" for Z-score method or "iqr" for IQR method. Inputted values are case insensitive. Named "method" to clearly indicate the desired approach to determine outliers.
#' @param zscore_threshold Numeric value equal to or greater than 1. The Z-score threshold (i.e number of standard deviations) above which values are considered outliers. Default is 3. Named "zscore_threshold" to specify the cut off for identifying outliers based on the Z-score method.
#' @param iqr_threshold Numeric value greater than 0. The multiplier for the IQR to define the bounds for detecting outliers. Default is 1.5. Named "iqr_threshold" to specify the multiplier for identifying outliers based on the IQR method. 
#' @param quantile_type Numeric integer (1-9). The algorithm used to compute quantiles when calculating the IQR. Default is 7. Please see quantile() documentation from the stats package for details. Named "quantile_type" to refer to the specific quantile algorithm chosen for the quantile() function based on input.
#'
#' @return A numeric vector of outlier values. If no outliers are found, it returns a message "No outliers found" and an empty numeric vector.
#'



```

## Exercise 3: Include examples (15 points)

For these examples, we are going to create numeric vectors and test the function on existing datasets.

### Example 1 - Simple vector with outliers for both methods

To start off, I'll demonstrate how this function can detect clear outliers using both methods:

```{r}
test_1 <- c(1:20, -50000)

check_outliers(test_1, "iqr")
check_outliers(test_1, "zscore")

```

### Example 2 - Simple vector with no outliers

In this example, no values qualify as outliers, demonstrating how the function returns an empty vector and message when none are present:

```{r}
test_2 <-c(1:10)

check_outliers(test_2, "iqr")
check_outliers(test_2, "zscore")

```

### Example 3 - Simple vector with outliers for one method, not the other

Although both methods detected the outlier in example 1, there are different underlying calculations between them, sometimes leading to cases where one method may identify values differently than the other. In this example, two apparent outliers are only identified by the IQR method because of the way it is calculated:

```{r}
test_3 <- c(0:10, -300, 500)

check_outliers(test_3, "iqr")


# However, if that same vector is checked using the Z-score method...
check_outliers(test_3, "zscore")
```

Although they seem like obvious outliers, if you check the calculations both values are <3 SD from the mean and therefore not outliers.

### Example 4 - Adjusting the threshold values

Let's try lowering the Z-score threshold from 3 to 1.5 SDs and see how that changes the sensitivity for detection:

```{r}
check_outliers(test_3, "zscore", zscore_threshold = 1.5)
```

The function now returns a vector with two values.



### Example 5 - Checking a vector with NA values

Next, let's try a vector with missing (NA) values. As mentioned in the documentation, the function should automatically omit NA values from the calculation, ensuring it does not interfere with the underlying mean, SD or quantile functions.

```{r}
test_4 <- c(1:10, NA, 10000)
check_outliers(test_4, "iqr")
check_outliers(test_4, "zscore")

```

Inclusion of the NA value in the inputted vector did not disrupt the output.


### Example 6 - Large vector with outliers

So far all the examples have been for smaller vectors, but how does it perform with more realistic, larger data? Lets create a vector with multiple clear outliers and see if they are detected (along with the ones generated in the vector):

```{r}
test_5 <- rnorm(2000, mean = 40, sd = 15)
test_5 <- c(test_5, -800, 1200)

check_outliers(test_5, "iqr")
check_outliers(test_5, "zscore")

```

Both methods are able to detect the outliers, with the IQR method detected additional ones.

### Example 7 - Real dataset

Finally, lets test how the function works in a real dataset. I have chosen the `gapminder` dataset for this example:

```{r}
library(gapminder)
head(gapminder)

# Lets check for outliers within the population variable, 'pop'

check_outliers(gapminder$pop, "iqr")
check_outliers(gapminder$pop, "zscore")

```

We get a vector of outliers for both methods, with IQR again being more sensitive.

## Exercise 4: Test the Function (25 points)

```{r}

# Load the testthat package
library(testthat)

# 1) For these tests, we should expect to return an empty numeric vector and a message for both methods
test_that("Vector with no NA and no outliers", {
  test_vec1 <- c(1:5)
  expect_equal(check_outliers(test_vec1, "zscore"), numeric(0)) 
  expect_message(check_outliers(test_vec1, "zscore"), "No outliers found.") 
  expect_equal(check_outliers(test_vec1, "iqr"), numeric(0))
  expect_message(check_outliers(test_vec1, "iqr"), "No outliers found.")
})


# 2) The NAs within the vector should not interfere with the return since they are being omitted, and return 5000 for both methods
test_that("Vector with NAs and outliers", {
  test_vec2 <- c(1:10, NA, 11:16, NA, 5000)
  expect_equal(check_outliers(test_vec2, "zscore"), 5000)
  expect_equal(check_outliers(test_vec2, "iqr"), 5000)
  
  # Lets also test that the tolower() function is working properly
  expect_equal(check_outliers(test_vec2, "ZsCoRe"), 5000)
  expect_equal(check_outliers(test_vec2, "IQr"), 5000)
  
})

# 3) These should all return errors as they are not numeric vectors greater than length 0
test_that("Non-numeric or >0 length vectors", {
  test_vec3 <- c("a", "b", "c")
  test_vec4 <- c(TRUE, FALSE, FALSE)
  test_vec5 <- c(1, 2, "s")
  expect_error(check_outliers(test_vec3, "zscore"))
  expect_error(check_outliers(test_vec3, "iqr"))
  expect_error(check_outliers(test_vec4, "zscore"))
  expect_error(check_outliers(test_vec4, "iqr"))
  expect_error(check_outliers(test_vec5, "zscore"))
  expect_error(check_outliers(test_vec5, "iqr"))
  expect_error(check_outliers(numeric(0), "zscore"))
  expect_error(check_outliers(numeric(0), "iqr"))
})


# 4) Test other miscellaneous errors and stops based on input ranges

test_that("invalid inputs or other messages", {
  
  # Test for when SD of a vector is 0 using the zscore method
  test_vec6 <- c(5, 5, 5, 5, 5)
  test_vec2 <- c(1:10, NA, 11:16, NA, 5000)
  expect_message(check_outliers(test_vec6, "zscore"), "Standard deviation is zero or NA; no outliers can be determined using this method.")
  expect_equal(check_outliers(test_vec6, "zscore"), numeric(0))  # Expect empty numeric vector
  
  
  # Test for when the z-score threshold is a non-numeric or less than 1
  expect_error(check_outliers(test_vec2, "zscore", zscore_threshold = TRUE), "The z-score threshold must be a numeric value at least 1 SD or greater.")
  expect_error(check_outliers(test_vec2, "zscore", zscore_threshold = 0.7), "The z-score threshold must be a numeric value at least 1 SD or greater.")
  
  # However, this should not interfere if the chosen method is not "zscore"
  expect_equal(check_outliers(test_vec2, "iqr", zscore_threshold = 0.2), 5000)
  
  # Test for when the IQR threshold multipler is a non-numeric or negative number
  expect_error(check_outliers(test_vec2, "iqr", iqr_threshold = -5), "The IQR threshold must be a numeric value greater than or equal to 0.")
 expect_error(check_outliers(test_vec2, "iqr", iqr_threshold = "7"), "The IQR threshold must be a numeric value greater than or equal to 0.")
 
 
 # Test for when IQR quantile algorithm is a non-integer between 1-9
  expect_error(check_outliers(test_vec2, "iqr", iqr_threshold = 1, quantile_type = 4.5), "Invalid quantile algorithm type; must be an integer between 1 and 9.")
 expect_error(check_outliers(test_vec2, "iqr", quantile_type = -3), "Invalid quantile algorithm type; must be an integer between 1 and 9.")
 
 # However, this should not interfere if the chosen method is not "iqr"
  expect_equal(check_outliers(test_vec2, "zscore", iqr_threshold = 0.2, quantile_type = 50), 5000)
 
  
})



```